\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-04 paper/paper.tex
version sent to scott

The media quickly picked up the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. While
many believed the moment was a reflection of the Olympic spirit, not
everyone agreed with this rosy interpretation. The twins' happy facial
expressions at the finish were portrayed as a bit contrived---smiling
like ``Honigkuchenpferde,'' cookies in the shape of a horse, was the
description offered by one editorialist---and the sports director of
the German Athletics Federation, Thomas Kurschilgen, stirred up
controversy when he suggested that the Hahners' photo-finish was no
coincidence.  Kurschilgen averred that the twins slowed down so as to
finish simultaneously and create a spectacle which would ``generate
media attention.''  Kurschilgen justified his charge with the fact
that the twins ran the Rio marathon over 18 minutes slower their
personal best times prior to the Olympics.  Not surprisingly,
Kurschilgen's accusations were denied by the Hahner twins, who claimed
that their simultaneous finish was simply an unintended coincidence.

@@

What happened in the women's Olympic marathon, and how might we
develop a statistical approach that assesses whether the Hahner twin's
finish was coincidental or intentional?  These two interpretations are
clearly at odds. If the former, then the Hahners are to be celebrated
and their finish treated as an expression of the spirit behind the
Olympic games. If the latter, though, then the twins may have violated
this spirt by not trying hard enough. It is perhaps too easy for us to
write such a glib sentence---neither of us can fathom being able to
complete a marathon anywhere in the vicinity of two and a half
hours---but we nonetheless want to know what the data from the Olympic
marathon tell us.

@@

The Kurschilgen accusation against the Hahner twins has two
components, that these two women ran slowly \emph{and} that they
finished simultaneously.  We suspect that Kurschilgen would not have
expressed ire at the Hahners had they finished in 1st and 2nd place in
Rio, hand-in-hand with wide grins, but of course we cannot be certain
of this.  Thus, our investigation of the charges that Kurschilgen
offered will distinguish between a slow finish versus a simultaneous
finish.

Our research design is twofold.  First, we present visualizations that
describe various features of the Hahner twin's results, and an
important element in our visualizations is the difference between
runners' Rio times and their prior personal best marathon times.  Our
visualizations suggest that the Hahner twin's pace in the marathon was
slow, albeit not excessively so, but that their simultaneous finish
was quite unusual given the twins' differences in abilities (and
similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, and our simulations reinforce what we
observed in prior visualizations, namely, that the Hahner twins did
not run appreciably slowly yet finished suspiciously close to each
other.  We return in our conclusion to Kurschilgen's claims about the
Hahners and offer thoughts about their validity.

@@

One way that we might assess whether an athlete's Rio finish was
unusual---or, say, whether two athletes' finishes were jointly
unusual---is by comparing a runner's observed Olympic women's marathon
finishing time on August 14, 2016, to a measure of her underlying
marathon talent.  For the latter we use prior personal best marathon
times.  This is a natural measure of marathon talent, we believe, but
it is not entirely free of complications.  For example, personal best
times are potentially confounded by the marathon courses on which they
were set; some courses, like the Berlin marathon, are known for
relatively fast times.  In addition, personal best times may be
confounded by conditions, like weather, that varied across races.
Finally, personal best times may not capture Olympic raceday
idiosyncrasies that could affect individual runners.  For example, a
runner could have woken up with a minor cold on the morning of August
14, 2016.  With these concerns in mind, we use a runner's half
marathon split time from the Olympic marathon as a secondary measure
of athletic skill in marathoning.  Our two measures of marathon talent
are featured in the visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain second-order polynomial regression lines.  The ticks along the
x-axis in both plots indicate times (either personal best or half
marathon split) of runners who earned  DNF results.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.

Consider first the Hahner twins.  They are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's orange dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

Thinking about the accusations leveled against the Hahner twins,
Figure \ref{fig:diffdiffscatter} suggests that Anna and Lisa Hahner
did indeed run with an eye on each other.  In fact, the same can be
said of the Kim twins, who ran seemingly in lockstep throughout the
entire Rio marathon.  The North Korean twins had a personal best
difference of 53 seconds and a finishing time difference of literally
zero seconds.  Beyond these twins, there were eight pairs of Rio
runners with a 53 second personal best difference, and resulting
finishing time differences are as follows: 9, 51, 228, 340, 352, 571,
662, and 751.  As in the Hahner case, the Kim twins compressed their
finishing times - meaning that they finished with less time between  
them than the difference in their personal bests - more than any other pair of runners with a similar
personal best difference.

@@

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to the accumulation of DNFs.  The usual color scheme
applies here, and the Estonian triplet DNF prior to the half marathon
split is evident. As noted in the figure Lisa Hahner was ahead of her
sister through 15 kilometers at which point Anna took the lead.
Figure \ref{fig:secondsbehind} contains two red dots representing the
North Korean Kim twins, but this is not visually apparent because the
Kim twins had identical split times during the entire marathon.

@@

Our visualizations are useful and they shed a fair bit of light on the
Hahner twin's performance in Rio.  In the interest of increasing
precision, we now pose the following question: if we take into
consideration the Hahner twin's similarities in marathon talent as
well as natural variation in marathon finishing times, what is the
probability that the twins would finish the Rio race at roughly the
same time and/or sequentially?  To answer this question, we need to
know the counterfactual distribution of potential marathon finishing
times that would have occurred had the Hahner twins independently (in
particular, of each other) and repeatedly run the Rio marathon,
holding constant marathon conditions, the abilities of other runners,
and so forth.  Access to such a distribution would establish the set
of potential outcomes that could have occurred on August 14, 2016, and
we could in principle use this distribution to determine the
likelihood that a simultaneous finish by the Hahners, or at least a
near-simultaneous finish, occurred by chance alone.  If these twins
rarely finish the marathon together in such a counterfactual world,
then one might be skeptical that their observed finish occurred
without coordination---as they in fact claimed.

@@

The histograms in Figure \ref{fig:simdiff} raise questions about the
credibility of Anna and Lisa Hahner's story. In particular, the
histogram suggest that a simultaneous finish in the Rio marathon would
be very rare if Anna and Lisa had run independently of each other. For
example, in fewer than 300 of the 10,000 simulated races did Anna and
Lisa finish within 30 seconds of each other. Moreover, they finished
in consecutive rank in fewer than 175 of the 10,000 simulated races.
The close finish that was observed, where Anna and Lisa crossed the
finish line one after the other, would have been highly unlikely if
the two German twins had raced independently of each other.

Parallel with our prior analyses, we repeated our simulations but this
time used half marathon splits as the predicting variable in our
simulations. The results of this are in Figure \ref{fig:simdiffhalf}.
While this use of marathon splits reduces the variation in the
predicted outcome of each runner and therefore reduces the expected
distance between Anna and Lisa Hahner, even with half marathon split
as a measure of ability it is still quite rare for Anna and Lisa to
finish simultaneously or consecutively.

@@

Our goal is not to speak to the question of whether the Hahner twins
should or should not have enjoyed a somewhat contrived moment of
Olympic glory. Neither was in contention for a podium finish in Rio,
and compared to the doping allegations that presently surround
endurance sports in general, what the Hahners did seems relatively
tame. Still, it might have behooved them to be a bit more open about
their end-of-race tactics, and we have shown here how a simple data
analysis can shed light on claims about racing results.  Should the
2020 Summer Olympics feature twins and triplets again, we look forward
to a comparative analysis using the techniques illustrated here.

>>>>

The media quickly picked up the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. While
many believed the moment was a reflection of the Olympic spirit, not
everyone agreed with this rosy interpretation. The twins' happy facial
expressions at the finish were portrayed as a bit contrived---smiling
like ``Honigkuchenpferde,'' cookies in the shape of a horse, was the
description offered by one editorialist---and the sports director of
the German Athletics Federation, Thomas Kurschilgen, stirred up
controversy when he suggested that the Hahners' photo-finish was no
coincidence.  Kurschilgen averred that the twins slowed down so as to
finish simultaneously and create a spectacle which would ``generate
media attention.''  Kurschilgen justified his charge with the fact
that the twins ran the Rio marathon over 18 minutes slower their
personal best times prior to the Olympics.  Kurschilgen's accusations
were denied by the Hahner twins, perhaps not surprisingly, who claimed
that their simultaneous finish was simply an unintended coincidence.

@@

What happened in the women's Olympic marathon in Rio, and how might we
develop a statistical approach that assesses whether the Hahner twin's
finish in the race was coincidental or intentional?  These two
interpretations are clearly at odds. If the former, then the Hahners
are to be celebrated and their finish treated as an expression of the
spirit behind the Olympic games. If the latter, though, then the twins
may have violated this spirt by not trying hard enough. It is perhaps
too easy for us to write such a glib sentence---neither of us can
fathom being able to complete a marathon anywhere in the vicinity of
two and a half hours---but we nonetheless want to know what the data
from the Olympic marathon tell us.

@@

Kurschilgen's accusation against the Hahner twins has two components,
that these two women ran slowly \emph{and} that they finished
simultaneously.  We suspect that Kurschilgen would not have expressed
ire at the Hahners had they finished in 1st and 2nd place in Rio,
hand-in-hand with wide grins.  Thus, our investigation of the charges
that Kurschilgen offered will distinguish between a slow finish versus
a simultaneous finish.

Our research design is twofold.  First, we present visualizations that
describe various features of the 2016 women's Olympic marathon, and an
important element in our visualizations is the difference between
runners' Rio times and their prior personal best marathon times.  Our
visualizations suggest that the Hahner twin's pace in the marathon was
slow, albeit not excessively so, but that their simultaneous finish
was quite unusual given the twins' differences in abilities (and
similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, and our simulations reinforce what we
observed in prior visualizations, namely, that the Hahner twins did
not run appreciably slowly yet finished suspiciously close to each
other.  We return in our conclusion to Kurschilgen's claims about the
Hahners and offer thoughts about their validity.

@@

One way that we might assess whether an athlete's Rio finish was
unusual---or, say, whether two athletes' finishes were jointly
unusual---is by comparing a runner's observed Olympic women's marathon
finishing time on August 14, 2016, with a measure of her underlying
marathon talent.  For the latter we use prior personal best marathon
times.  This is a natural measure of marathon talent, we believe, but
it is not entirely free of complications.  For example, personal best
times are potentially confounded by the marathon courses on which they
were set; some courses, like the Berlin marathon, are known for
relatively fast times.  In addition, personal best times may be
confounded by conditions, like weather, that varied across races.
Finally, personal best times may not capture Olympic raceday
idiosyncrasies that could affect individual runners.  For example, a
runner could have woken up with a minor cold on the morning of August
14, 2016.  With these concerns in mind, we use a runner's half
marathon split time from the Olympic marathon as a secondary measure
of athletic skill in marathoning.  Our two measures of marathon talent
are featured in the visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain second-order polynomial regression lines.  The ticks along the
horizontal axes in both plots indicate times, either personal bests or
half marathon splits, of runners who earned DNF results.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's orange dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

Thinking about the accusations leveled against the Hahner twins,
Figure \ref{fig:diffdiffscatter} suggests that Anna and Lisa Hahner
did indeed run with an eye on each other.  In fact, the same can be
said of the Kim twins, who ran seemingly in lockstep throughout the
entire Rio marathon.  The North Korean twins had a personal best
difference of 53 seconds and a finishing time difference of literally
zero seconds.  Beyond these twins, there were eight pairs of Rio
runners with a 53 second personal best difference, and resulting
finishing time differences are as follows: 9, 51, 228, 340, 352, 571,
662, and 751.  As in the Hahner case, the Kim twins compressed their
finishing times---meaning that they finished with less time between
them than the difference in their personal bests---more than any other
pair of runners with similar personal best differences.

@@

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to the accumulation of DNFs.  The usual color scheme
applies here, and the Estonian triplet DNF prior to the half marathon
split is evident. As noted in the figure, Lisa Hahner was ahead of her
sister through 15 kilometers at which point Anna took the lead.
Figure \ref{fig:secondsbehind} contains two red dots representing the
North Korean Kim twins, but this is not visually apparent because the
Kim twins had identical split times during the entire marathon.

@@

Our visualizations are useful and shed a fair bit of light on the
Hahner twin's performance in the 2016 women's Olympic marathon.  In
the interest of increasing precision, we now pose the following
question: if we take into consideration the Hahner twin's similarities
in marathon talent as well as natural variation in marathon finishing
times, what is the probability that the twins would finish the Rio
race at roughly the same time and/or sequentially?  To answer this
question, we need to know the counterfactual distribution of potential
marathon finishing times that would have occurred had the Hahner twins
independently (in particular, of each other) and repeatedly run the
Rio marathon, holding constant marathon conditions, the abilities of
other runners, and so forth.  Access to such a distribution would
establish the set of potential outcomes that could have occurred on
August 14, 2016, and we could in principle use this distribution to
determine the likelihood that a simultaneous finish by the Hahners, or
at least a near-simultaneous finish, occurred by chance alone.  If
these twins rarely finish the marathon together in such a
counterfactual world, then one might be skeptical that their observed
finish occurred without coordination.

@@

The histograms in Figure \ref{fig:simdiff} raise questions about the
credibility of Anna and Lisa Hahner's story and in particular suggest
that a simultaneous finish in the Rio marathon would be very rare if
Anna and Lisa had run independently of each other. For example, in
fewer than 300 of the 10,000 simulated races did Anna and Lisa Hahner
finish within 30 seconds of one another. Moreover, they finished in
consecutive rank in fewer than 175 of the 10,000 simulated races.  The
close finish that was observed, where Anna and Lisa crossed the finish
line one after the other, would have been highly unlikely if the two
German twins had raced independently of each other.

Parallel with our prior analyses, we repeated our simulations but this
time used half marathon splits as the predicting variable in our
simulations. The results of this are in Figure \ref{fig:simdiffhalf}.
While this use of marathon splits reduces the variation in the
predicted outcome of each runner and therefore reduces the expected
distance between Anna and Lisa Hahner, even with half marathon split
as a measure of ability it is still quite rare for the twins to finish
simultaneously or consecutively.

@@

Our goal is not to speak to the question of whether the Hahner twins
should or should not have enjoyed a somewhat contrived moment of
Olympic glory. Neither was in contention for a podium finish in Rio,
and compared to the doping allegations that presently surround
endurance sports in general, what the Hahners did seems relatively
tame. Still, it might have behooved them to be a bit more open about
their end-of-race tactics, and to this end we have shown here how a
simple data analysis can shed light on claims about racing results.
Should the 2020 Summer Olympics feature twins and triplets again, we
look forward to a comparative analysis using the techniques
illustrated here.
